---
title: "Lab 07 — t-tests and ANOVA"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab extends hypothesis testing from **claims about one mean** to:

- **one-sample t-tests**
- **two-sample t-tests**
- **paired t-tests**
- **ANOVA**

You will use the **US-AMS simplified AmeriFlux dataset** again and compare means across different time-based groups.

The core logic stays the same:

> estimate → uncertainty → test statistic → p-value

This lab emphasizes both **manual understanding** and **automated testing in R**.

---

# Learning goals

By the end of this lab, you should be able to:

- Compute and interpret a **one-sample t-test**
- Distinguish between **two-sample** and **paired** t-tests
- Compute and interpret **degrees of freedom**
- Explain the structure of an **ANOVA**
- Compare manual calculations to `t.test()` and `aov()`
- Use visualizations to support statistical interpretation
- Produce a fully reproducible R Markdown analysis

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=FALSE}
# GOAL: Read the dataset and keep 2022–2023.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local %in% c(2022, 2023))

glimpse(df)
```

## Derive time columns if needed

```{r derive_time, echo=TRUE, eval=FALSE}
# GOAL: Create date / month columns if they do not already exist.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE)
  )
head(df$doy)
tail(df$doy)
```

---

# Part I — One-sample t-test

## 1) Choose a day and a variable

Select:
- one **year**
- one **numeric variable**

Examples:
- `ta`
- `swc`
- `le`
- `fc`

```{r choose_one_sample, echo=TRUE, eval=FALSE}
target_year <- 2023
target_var  <- "ta"

sample_data <- df %>%
  filter(year_local == target_year) %>% 
  slice_sample(n = 200)

x <- sample_data[[target_var]]

mean(x, na.rm = TRUE); sd(x, na.rm = TRUE); length(x)
```

> What variable and day did you choose, and why?
I chose air temperature in 2023 to continue the explore this variable from the previous labs.
---

## 2) State a null hypothesis

Write a claim about the mean of your chosen variable on that day.

Examples:
- “The mean temperature on this day is 15 °C.”
- “The mean latent heat flux on this day is 100 W m$^{-2}$.”

Fill in:

> $H_0$: `The mean temperature on this day is 13 °C`

> $H_A$: `The mean temperature on this day is greater than 13 °C`

---

## 3) Compute the one-sample t-statistic manually

Use:

$t = (\bar{x} - \mu_0) / (s / \sqrt{n})$

```{r one_sample_manual, echo=TRUE, eval=FALSE}
mu0 <- 13

xbar <- mean(x, na.rm = TRUE)
s <- sd(x, na.rm = TRUE)
n <- sum(!is.na(x))

t_stat <- (xbar - mu0) / (s / sqrt(n))
t_stat
```

---

## 4) Compute the p-value and degrees of freedom

```{r one_sample_pvalue, echo=TRUE, eval=FALSE}
df_t <- x - 1

p_value <- 2 * pt(-abs(t_stat), df = df_t)

tibble(
  xbar = xbar,
  s = s,
  n = n,
  df_t = df_t,
  t_stat = t_stat,
  p_value = p_value
)
```

> What does the p-value suggest about your null hypothesis?
The p-value is about 0.9, which is greater than 0.05 significance level, therefore failing to reject the null hypothesis. 


---

## 5) Compare with `t.test()`

```{r one_sample_ttest, echo=TRUE, eval=FALSE}
t.test(x, mu = mu0)
```

> Does the `t.test()` output agree with your manual calculation? If not, why might that be?
The t-test output does not have the same t-statistic as the manual calcualtion, however, it does share the same p-value and mean. The manual calculation has a t-value of 1.36, while this t-test is 0.12. This may be because of variations in the input data when calculating the t-stat. 
---

# Part II — Two-sample t-test

## 6) Compare the same variable across two day ranges

Choose:
- the **same variable**
- two different **day ranges in 2023**

```{r choose_two_sample, echo=TRUE, eval=FALSE}
x1 <- df %>%
  filter(year_local == 2023, doy > 144 & doy < 244) %>%
  pull(target_var)

x2 <- df %>%
  filter(year_local == 2023, doy > 244 & doy < 354) %>%
  pull(target_var)

mean(x1, na.rm = TRUE); sd(x1, na.rm = TRUE); length(x1)
mean(x2, na.rm = TRUE); sd(x2, na.rm = TRUE); length(x2)
```

> Why did you choose these two days?
I wanted to see the full range of the available days which goes from about 144 to 354. 
---

## 7) Summarize the difference in means

```{r two_sample_summary, echo=TRUE, eval=FALSE}
xbar1 <- mean(x1, na.rm = TRUE)
xbar2 <- mean(x2, na.rm = TRUE)

s1 <- sd(x1, na.rm = TRUE)
s2 <- sd(x2, na.rm = TRUE)

n1 <- sum(!is.na(x1))
n2 <- sum(!is.na(x2))

mean_diff <- xbar1 - xbar2

tibble(
  xbar1 = xbar1,
  xbar2 = xbar2,
  mean_diff = mean_diff
)
```

---

## 8) Compute the two-sample t-statistic manually (Welch)

```{r two_sample_manual, echo=TRUE, eval=FALSE}
se_diff <- sqrt((s1^2 / n1) + (s2^2 / n2))

t_stat <- mean_diff / se_diff

df_welch <- ((s1^2 / n1 + s2^2 / n2)^2) /
  (((s1^2 / n1)^2 / (n1 - 1)) + ((s2^2 / n2)^2 / (n2 - 1)))

p_value <- 2 * pt(-abs(t_stat), df = df_welch)

tibble(
  mean_diff = mean_diff,
  se_diff = se_diff,
  t_stat = t_stat,
  df_welch = df_welch,
  p_value = p_value
)
```

> What does the sign of `mean_diff` mean? What does the p-value mean?
The sign of the mean_diff is positive meaning the first mean calculated is higher than the second mean. The p-value is 0 which is evidence against the null hypothesis. 

## 9) Compare with `t.test()`

```{r two_sample_ttest, echo=TRUE, eval=FALSE}
t.test(x1, x2)
```

> Does the automated t-test agree with your manual result?
The automated t-test does agree with the manual result, with both t-statistics being 77.5. 
---

## 10) Visualize the two groups

```{r two_sample_plot, echo=TRUE, eval=FALSE}
plot_df <- bind_rows(
  tibble(group = paste0("DOY_"), value = x1),
  tibble(group = paste0("DOY_"), value = x2)
)
#group = paste0("DOY_", day_range1)
#group = paste0("DOY_", day_range2),
ggplot(plot_df, aes(x = group, y = value, fill = group)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = target_var,
    title = "Two-sample comparison"
  )
```

> Does the plot match your statistical conclusion?
This plot does match the statistical conclusion because it shows a small spread and a large outlier that may contribute to the randomness that was concluded. 
---

# Part III — Paired t-test

## 11) Compare the same days across two years

Choose one DOY and compare **2022 vs 2023** for the same variable.

```{r paired_setup, echo=TRUE, eval=FALSE}
x1 <- df %>%
  filter(year_local == 2022, doy > 144 & doy < 244) %>%
  arrange(date) %>%
  pull(target_var)

x2 <- df %>%
  filter(year_local == 2023, doy > 144 & doy < 244) %>%
  arrange(date) %>%
  pull(target_var)

length(x1); length(x2)
```

> Why is this a paired comparison rather than an independent two-sample comparison?
This is paired because it compares the same time frame between the two years. 
---

## 12) Compute paired differences manually

```{r paired_manual, echo=TRUE, eval=FALSE}
d <- x1 - x2

dbar <- mean(d, na.rm = TRUE)
sd_d <- sd(d, na.rm = TRUE)
n <- sum(!is.na(d))

t_stat <- dbar / (sd_d / sqrt(n))
p_value <- 2 * pt(-abs(t_stat), df = n - 1)

tibble(
  dbar = dbar,
  sd_d = sd_d,
  n = n,
  t_stat = t_stat,
  p_value = p_value
)
```

> What does the sign of `dbar` mean?
The dbar is the mean of the difference of the two variables. A postive dbar means that the values in the x2 are larger than x1 on average.
---

## 13) Compare with `t.test(..., paired = TRUE)`

```{r paired_ttest, echo=TRUE, eval=FALSE}
t.test(x1, x2, paired = FALSE)
```

> How does the paired t-test differ conceptually from the two-sample t-test?
For these two test the p-values are similar but the t-statistics are not. Conceptually, the paired t-test samples are related in some way while the two-sample t-test are independent of eachother.
---

# Part IV — ANOVA

## 14) Create a categorical grouping variable

Group `ta` into broad seasonal categories:

- Summer
- Winter
- Other

```{r make_groups_anova, echo=TRUE, eval=FALSE}
df_anova <- df %>%
  filter(!is.na(ta)) %>%
  mutate(
    month_group = case_when(
      month %in% c(20, 22, 23) ~ "Summer",
      month %in% c(10, 11, 13) ~ "Winter",
      TRUE ~ "Other"
    )
  )

df_anova %>% count(month_group)
```

---

## 15) Visualize the groups

```{r plot_anova_groups, echo=TRUE, eval=FALSE}
ggplot(df_anova, aes(x = month_group, y = ta, fill = month_group)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = "ta",
    title = "Air temperature by seasonal group"
  )
```

> Do the groups look different enough that ANOVA might detect a difference?
Based on this boxplot, the groups do not show much of a difference between eachother, so the ANOVA might not detect a difference. 
---

## 16) Compute ANOVA components manually

```{r anova_manual_components, echo=TRUE, eval=FALSE}
grand_mean <- mean(df_anova$ta, na.rm = TRUE)

group_stats <- df_anova %>%
  group_by(month_group) %>%
  summarise(
    group_mean = mean(ta, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

ss_between <- sum(group_stats$n * (group_stats$group_mean - grand_mean)^2)

df_anova2 <- df_anova %>%
  left_join(group_stats, by = "month_group")

ss_within <- sum((df_anova2$ta - df_anova2$group_mean)^2)

k <- n_distinct(df_anova$month_group)
n_total <- nrow(df_anova)

df_between <- k - 1
df_within <- n_total - k

ms_between <- ss_between / df_between
ms_within <- ss_within / df_within

f_stat <- ms_between / ms_within
p_value <- pf(f_stat, df1 = df_between, df2 = df_within, lower.tail = FALSE)

tibble(
  grand_mean = grand_mean,
  ss_between = ss_between,
  ss_within = ss_within,
  df_between = df_between,
  df_within = df_within,
  f_stat = f_stat,
  p_value = p_value
)
```

> In your own words, what do `ss_between` and `ss_within` represent?
The ss_between represents the mean differences between the groups while the ss_within are the differences within the groups. 
---

## 17) Compare with `aov()`

```{r run_anova, echo=TRUE, eval=FALSE}
model <- aov(ta ~ month_group, data = df_anova)
summary(model)

```

> What does the ANOVA p-value tell you? Does it tell you which groups differ?
The ANOVA did not compute, perhaps because the ANOVA could not differ between the varying groups. 
---

## 18) Final reflection

Answer in **5–7 sentences**:

- How is a one-sample t-test different from a two-sample t-test?
- Why is a paired t-test special?
- How is ANOVA a generalization of the t-test?
- What stayed the same across all these tests?

> 
A one-sample t-test creates an estimate for a claim about a population parameter, while a two-sample t-test comapares two different population parameter estimates. A paired t-test is special because the  two variables are related to eachother in someway, being dependent of eachother. An ANOVA is a generalization of the t-test because it can compare more than two groups, which means is assumes more in its calculations. Across all four tests, the means stayed about the same. This is likely because they are pulling from the same sample.  
---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
- Submit the GitHub repository link on Canvas.
